BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2014-11-21 Paper/BOAST.tex
add first draft of SPECFEM use case
Forst modifications in the paper for BigDFT

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{listings}
\usepackage{color}

@@

\usepackage{url}
\modulolinenumbers[5]

@@

\linenumbers

@@

  \subsection{Creating an Auto-Tuned Convolution Library for BigDFT using BOAST}

@@

BigDFT~\cite{bigdft} is a scientific application computing the electronic
structure of systems using density functional
theory (DFT). The application is mainly written in Fortran and currently
includes some 193,000 lines of Fortran code, which accounts for more than 50 \% 
of the code base. It is a parallel application based on the standards
MPI~\cite{mpi} and OpenMP~\cite{openmp}.  It also supports CUDA~\cite{cuda} and
OpenCL~\cite{opencl}.  As it is used on systems that may
have very different architectures, it is of great importance to be able to
optimize and run the application according to the specific underlying platform.

BigDFT is characterized by its extensive use of
convolution operators~\cite{nussbaumer1982fast} applied on large arrays of
data.  One example of a specific convolution, called
MagicFilter~\cite{Genovese|CAS2010}, can be seen in Listing~\ref{lst:conv_example}. 
It applies a filter $filt$ to the data set $in$ and then stores the result in the data set $out$ with a
transposition~\cite{Goedecker1993}.

@@

As we can see, there are three nested loops working on arrays whose sizes
vary.  Various optimizations can be applied to this treatment and
may focus on the loop structure, as well as on the size of the data.  In this
paper we focus on this particular convolution and explore different
optimizations whose nature is presented in detail in the next section.

    \subsubsection{A Generic Convolution Library}

  BigDFT implementation of convolution/wavelets is a reference but is limited in scope (1 wavelet family ands its derived operators).
  Generalizing to other wavelet families long and difficult if performance are to remain good.

Idea:
\begin{itemize}
\item write a generic convolution library in BOAST
\item BLAS like interface
\item Provide a wide variety of building blocks
\item Optimize thos blocks using BOAST
\item Build either from source or directly link with binary
\item Release complete library
\end{itemize}

  Characteristic use case driven

  Convolution Library: first step toward a dsl dedicated to wavelet processing for Physics computing.

  \subsection{Porting SPECFEM3D to OpenCL using BOAST}

    \subsubsection{SPECFEM3D}

\begin{itemize}

\item Wave propagation research software

\item HPC application (benchmark on supercomputers)

\item Complex GPU (CUDA) code, 40+ kernels.

\item 80+ parameters for kernels. Difficult to validate.
\end{itemize}

    \subsubsection{Porting to OpenCL}

  Idea:
\begin{itemize}
\item write kernels in BOAST,
\item generate kernels in CUDA,
\item validate kernels by comparing CUDA/generated CUDA executions using comparator
\item generate kernels in OpenCL,
\item develop OpenCL runtime backend using previously validated OpenCL kernels
\item validate OpenCL version by comparing CUDA/OpenCL executions using comparator
\end{itemize}

  Evaluation:
\begin{itemize}
\item Similar CUDA/OpenCL
\item Factoring : more compact code
\item Integrated non regression support
\item Integrated Kernel performance evaluation
\end{itemize}

>>>>

\documentclass%[review]
              {elsarticle}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{color}
\usepackage{xspace}

%\usepackage{lineno}
%\modulolinenumbers[5]
%\linenumbers

@@

\usepackage{url}

@@

@@

  \subsection{Creating an Auto-Tuned Convolution Library for BigDFT using BOAST}
In 2005, the EU FP6-STREP-NEST BigDFT project funded a consortium of four European
laboratories (L\_Sim - CEA Grenoble, Basel University - Switzerland, Louvain-la-Neuve University - Belgium and
Kiel University - Germany), with the aim of developing a novel approach for DFT calculations
based on Daubechies wavelets.
Rather than simply building a DFT code from scratch, the objective of this three-years
project was to test the potential benefit of a new formalism in the context of electronic structure
calculations.

As a matter of fact, Daubechies wavelets exhibit a set of properties which make them ideal
for a precise and optimized DFT approach. In particular, their systematicity allows to provide a
reliable basis set for high-precision results, whereas their locality (both in real and reciprocal space)
is highly desired to improve the efficiency and the flexibility of the treatment. Indeed, a localized
basis set allows to optimize the number of degrees of freedom for a required accuracy (see \cite{bigdft}),
which is highly desirable given the complexity and inhomogeneity of the systems under investigation
nowadays.

The application is mainly written in Fortran and currently
includes about 360 thousands lines of Fortran code and 70 thousands lines of C language, which accounts for more than 50 \% 
of the code base. It is a parallel application based on the standards
MPI~\cite{mpi} and OpenMP~\cite{openmp}.  It also supports CUDA~\cite{cuda} and
OpenCL~\cite{opencl}.  
In the recent years this code has been used for many scientific applications,
and its development and user consortium is continuously growing.
Massively parallel computations are routinely executed with the BigDFT code, either in homogeneous or 
hybrid architectures. 
In 2009, the French Bull-Fourier was awarded for the implementation of the hybrid version of BigDFT.

%As it is used on systems that may
%have very different architectures, it is of great importance to be able to
%optimize and run the application according to the specific underlying platform.

@@

In the Kohn-Sham (KS) formulation of DFT, the electrons are associated to wavefunctions (orbitals), which are
represented arrays of floating point numbers. In wavelets formalism, the operators are written via convolutions with short,
separable filters.
The detailed description of how these operations are defined is beyond the scope of this paper and can be found
in the BigDFT reference paper \cite{bigdft}.

% Convolutions are among the basic processing blocks of BigDFT. Special care has to be
% taken regarding their performances. 
The CPU convolutions of BigDFT have thus been thoroughly optimized.
In a recent paper, the optimization of the CPU convolutions of BigDFT has been extensively considered.
%BigDFT is characterized by its extensive use of
%convolution operators~\cite{nussbaumer1982fast} applied on large arrays of
%data.  
One example of a specific convolution, called
MagicFilter~\cite{Genovese|CAS2010}, can be seen in Listing~\ref{lst:conv_example}. 
It applies a filter $filt$ to the data set $in$ and then stores the result in the data set $out$ with a
transposition~\cite{Goedecker1993}.

@@

As we can see, there are three nested loops working on arrays whose sizes
vary.  Various optimizations can be applied to this treatment and
may focus on the loop structure, as well as on the size of the data. 
%In this
%paper we focus on this particular convolution and explore different
%optimizations whose nature is presented in detail in the next section.

    \subsubsection{A Generic Convolution Library}
The number of convolution kernels needed in BigDFT has been continuously growing in the recent years.
Various boundary conditions and functionalities have been added, making the BigDFT more and more powerful
in terms of scientific applications.
However, the cost of the maintainance and of development of the convolutions is always a 
delicate point to be considered while including a new functionality.
The convolution patterns are usually rather similar, leading to code duplication 
and difficulties in code reading and modification.

It appears therefore very interesting to benefit from a automatic tool to drive the implementation and the 
generation of new convolutions.
This would lead to optimized code, adapted to  different computing platforms, that
is optimally factorized.
In addition to this point, the help of such code generator is also important to
build \emph{new science}: the cost of implementing new convolutions would become so little that other 
functionalities (for example, the generalisation of the BigDFT convolutions to Neumann boundary 
conditions  or the usage of wavelet-on-the-interval basis) can be added with limited manpower.

For these reasons, a convolution library has been engineered with the help of BOAST.
The detailed API of the library will be described elsewhere.
The spirit is similar to BLAS-LAPACK API, where low-level operations are scheduled and called 
from high-level operations.
Each of the building blocks of these convolution library is optimally tuned by BOAST by choosing
the sources providing the optimal kernel for the chosen computing platform.
The sources of these kernels are then collected and compiled to meet the API of the library.

A library written in this way might have an impact going largely beyond the community of BigDFT developers.
Indeed, convolutions are basic operations of lots of scientific application codes, like for example
finite differences approaches, which are universally used in computational physics.

The interest in having a robuts and optimally tuned library in this scientific field is therefore evident.
Techniques are under investigation to provide also fine-tuned binaries to the end-user rather than
the code sources, such that more agressive interprocedural optimisation can be performed.
This experience would be therefore a first step toward the release of a tunable optimised 
convolution library oriented to computational physics communities.
%   BigDFT implementation of convolution/wavelets is a reference but is limited in scope (1 wavelet family ands its derived operators).
%   Generalizing to other wavelet families long and difficult if performance are to remain good.
% 
% Idea:
% \begin{itemize}
% \item write a generic convolution library in BOAST
% \item BLAS like interface
% \item Provide a wide variety of building blocks
% \item Optimize thos blocks using BOAST
% \item Build either from source or directly link with binary
% \item Release complete library
% \end{itemize}
% 
% 
%   Characteristic use case driven
% 
%   Convolution Library: first step toward a dsl dedicated to wavelet processing for Physics computing.

\input{specfem}
